

## Simulation and model code for the Car-Markov tracking transect model. 

******

## Purpose:

R code to simulate and analyze tracking transect data from mulitple independent transects
with spatially replicated segments within transects and temporally replicated surveys. The 
method of inducing spatial correlaiton among transect segments, and much of the simulation
code, is derived from Aing et al. (2011).

Aing, C., S. Halls, K. Oken, R. Dobrow, and J. Fieberg. 2011. A Bayesian hierarchical occupancy model for track surveys conducted in a series of linear, spatially correlated, 
	sites. Journal of Applied Ecology 48:1508-1517.

Author: Andrew D. Crosby,
        Boreal Avian Modeling Project,
	      Department of Biological Sciences,
        University of Alberta,
        CW405 Biological Sciences Building,
        Edmonton, AB T6G 2E9

******

First of all, make sure you have WinBUGS installed on your computer with the key for unrestricted use and the update patch for version 1.4.3. Download the program and follow the instructions here: https://www.mrc-bsu.cam.ac.uk/software/bugs/the-bugs-project-winbugs/. 
I highly recommend using WinBUGS rather than OpenBUGS, because for some reason OpenBUGS does not handle the model consistently - in our trials it took much longer to run and did not recover the parameters as well as WinBUGS. We did not use JAGS becasue it does not have a built-in CAR prior. However, if you are skilled enough to write your own CAR prior for JAGS it might be a good idea to try it. 

Note: It is sometimes necessary to copy and paste the "WinBUGS14" from Program Files into your "Documents" folder because for some reason R wants administrator privalages to call it. In this case, you will need to specify the "bugs.directory" in the bugs command for running the model. 

The first set of code simulates csimulates the occupancy, use, track-laying, and observation processes. 

```{r}
rm(list=ls(all=TRUE))


ntrans <- 100  			# The number of transects 	            
nsegs<-10			# The number of segments per transect
nsurvs<-3			# The number of replicate surveys
acts.survs<-c(2, 4, 6)		# The number of activity periods after which each survey replicate was done	
max.act<-max(acts.survs)	# The maximum number of activity periods


# Simualte constant occupancy among all transects
psi<-0.6					# The probability of transect occupancy
occupancy<-rbinom(ntrans, 1, psi)	# The true occupancy state of each transect
mean(occupancy)

# Set the parameters of the segment use data
theta <- .3 		# The average probability of use among all transect segments
bo <-log(theta/(1-theta))	# The intercept for the use process
arparam <- 0.8		# The unconditional variance parameter for the ar(1) process
sdparam <- 2		# The autocorrelation parameter for the ar(1) process


# Simulate the probability of use at each segment|transect occupancy (theta[i]|z[i])
eps.theta<-matrix(nrow=ntrans, ncol=nsegs)
for(trans in 1:ntrans){
  eps.theta[trans,1:nsegs]<-as.numeric(arima.sim(n=nsegs, list(ar=arparam), sd=sdparam))
  }

theta.segs<-matrix(nrow=ntrans, ncol=nsegs)
for(trans in 1:ntrans){
  theta.segs[trans,]<-plogis(bo+eps.theta[trans,])*occupancy[trans]
}
theta.trans<-apply(theta.segs, 1, mean) 	# Mean segment use probability on each transect
theta.mean<-mean(theta.trans)		# Mean segment use probability over all transects
theta.mean


# Simulate the true use at each site 
usegs<-matrix(nrow=ntrans, ncol=nsegs)
for(trans in 1:ntrans){
 usegs[trans ,]<-rbinom(nsegs, 1, theta.segs[trans ,])
}
usegs.trans<-apply(usegs, 1, mean)	# Proportion of segments used on each transect
usegs.mean<-mean(usegs.trans)	# Mean segment use over all transects
usegs.mean



# Set the parameters of the track-laying and observation processes
deltatilde<-0.2		# The probability of a track being laid at the first segment (or mean probability of a track being laid)
beta<-0.5			# The increase in probability of a track being laid, given that the segment is occupied and a track was laid on the previous segment 

deltastar<-rep(NA, ntrans)				
for(trans in 1:ntrans){	
  deltastar[trans]<-deltatilde*(1-usegs.trans[trans]*beta)	# The probability of a track being laid , given that the segment is occupied and no track was laid at the previous segment 
  }
deltastar.mean<-mean(deltastar)

p<-0.95			# The probability of detecting a track given that it was laid



# Simulate the track-laying process

# Create the empty datasets
tracks<-array(NA, dim=c(ntrans, nsegs, max.act))	# The track-laying data
track_laid<-array(NA, dim=c(ntrans, nsegs, max.act))	# The cumulative track data
y.day<-array(NA, dim=c(ntrans, nsegs, nsurvs))        # The track data from each survey


  # Simulate the first segment of each transect
  for(trans in 1:ntrans){
    for(act in 1:max.act){
	tracks[trans,1,act]<-usegs[trans,1]*(1* (runif(1) < deltatilde)) # Track laid down?
	track_laid[trans,1,act]<-min(1, sum(tracks[trans,1,1:act]))
	}
  }

  # All remaining sites
for(trans in 1:ntrans){
  for(seg in 2:nsegs){
    for(act in 1:max.act){
	tracks[trans,seg,act]<-usegs[trans,seg]*(1*(runif(1)<(deltastar[trans]+tracks[trans,seg-1,act]*beta)))  # Track laid down?
	track_laid[trans,seg,act]<-min(1, sum(tracks[trans,seg,1:act]))
	}
  }
}

# Simulate the observation process for multiple consecutive surveys
y.day<-array(NA, dim=c(ntrans, nsegs, nsurvs))
for(trans in 1:ntrans){
  for(seg in 1:nsegs){
    for(surv in 1:nsurvs){
	y.day[trans,seg,surv]<-rbinom(1, 1, track_laid[trans,seg,acts.survs[surv]]*p)
    }
  }
}

# Summarize the data
occ.true<-mean(occupancy)
tracks.true<-mean(apply(tracks, 1, max))
use.obs<-mean(apply(y.day, c(1,2), max))
occ.obs<-mean(apply(apply(y.day, c(1,2), max), 1, max))


occ.true      # The true occupancy rate accross all transects
theta.mean    # The mean proabbility of segment use across all transects and segments
usegs.mean    # The true use rate across all transects and segements
tracks.true   # The mean number of transects where tracks were laid
use.obs       # The observed segement level use rate (naive estimate)
occ.obs       # The observed occupancy rate (naive estimate)

```


Next, create the neighborhood data required for the CAR model

```{r}
#  Create neighbor indices and adjacency matrix for CAR prior (Make sure there are at least 2 sites)
getNumNeighbors <- function(sites) {
    if (sites < 1)
        return(NULL)
    else if (sites == 1)
        return(c(0))
    else
        return(c(1, rep(2, sites - 2), 1))
}

getAdjMatrix <- function(sites) {
    if (sites < 2)
        return(NULL)
    else {
        adj_matrix <- rep(0, 2 * sites - 2)
        # First site's only neighbor is the second site.
        adj_matrix[1] = 2
        for (i in 2:(sites - 1)) {
            adj_matrix[2 * i - 2] = i + 1
            adj_matrix[2 * i - 1] = i - 1
        }
        # Last site's only neighbor is the second to last site.
        adj_matrix[length(adj_matrix)] = sites - 1
        return(adj_matrix)
    }
}

num <- getNumNeighbors(nsegs)
sum_neighbors <- sum(num)
adj <- getAdjMatrix(nsegs)

```

Create the CAR-Markov model for WinBUGS

```{r}

sink("model.txt")
cat("
model {

# Prior on occupancy
psi ~ dunif(0, 1)



# Priors on logit occupancy parameters
for(trans in 1:ntrans){
  alpha1[trans,1:nsegs] ~ car.normal(adj[], weights[], num[], tau1) # Site level random effects
  }
alpha0 ~ dflat() # Intercept
tau1 ~ dgamma(0.5, 0.0005) # conditional precision parameter (1/ variance)

for (i in 1:sum_neighbors) {
  weights[i] <- 1 
  }

# Prior for track-laying and detection
for(trans in 1:ntrans){
  deltastar[trans] ~ dunif(0, 1)
  }
beta ~ dunif(-0.1, 1)
p ~ dunif(0, 1)

# Model for the occupancy process
for(trans in 1:ntrans){
  z[trans]~dbern(psi)
  }

# Model for the segment use process
for(trans in 1:ntrans){
  for(seg in 1:nsegs){
    logit(theta[trans,seg])<-alpha1[trans,seg]+alpha0
    eta[trans,seg]<-z[trans]*min(0.999, max(0.001, theta[trans,seg]))
    u[trans,seg]~dbern(eta[trans,seg])
    }
  thetahat[trans]<-mean(theta[trans,1:nsegs])
  }

# Model for the track-laying process
  # for the first segment of each transect
  for(trans in 1:ntrans){
    const[trans]<-min(0.999, max(0.001,deltastar[trans]/(1-thetahat[trans]*beta)))
    for(act in 1:max.act){
	mu[trans,1,act]<-u[trans,1]*const[trans]		# Probability of track laid	
	tracks[trans,1,act]~dbern(mu[trans,1,act])	# Track present?
	w[trans,1,act]<-min(1, sum(tracks[trans,1,1:act]))	# Cumulative track index
	}
    }

# For all subsequent segments
for(trans in 1:ntrans){
  for(seg in 2:nsegs){
    for(act in 1:max.act){
	mu[trans,seg,act]<-u[trans,seg]*min(0.999, max(0.001,deltastar[trans]+tracks[trans,seg-1,act]*beta))
	tracks[trans,seg,act]~dbern(mu[trans,seg,act])	# Track present?
	w[trans,seg,act]<-min(1, sum(tracks[trans,seg, 1:act]))	# Cumulative track index
	}
    }
  }

# Model for the detection process
for(trans in 1:ntrans){
  for(seg in 1:nsegs){
    for(surv in 1:nsurvs){
	lam[trans,seg,surv]<-w[trans,seg,acts.survs[surv]]*p
	y.day[trans,seg,surv]~dbern(lam[trans,seg,surv])
	}
    }
  }

# Derived parameters
psihat<-mean(z[])
thetamean<-mean(thetahat[])
deltamean<-mean(const[])
deltastarmean<-mean(deltastar[])
uhat<-mean(u[,])

}

",fill=TRUE)

sink()

```

Finally, run the model on the simulated data

```{r}
# ----------------------------------------------------------------------------------
# Run the CAR-Markov model accounting for spatial correlation in theta and delta
# ----------------------------------------------------------------------------------

library(R2WinBUGS)

data<-list("y.day", "ntrans", "nsegs", "nsurvs", "max.act", "acts.survs", "sum_neighbors",
  "adj", "num")

inits<-function(){ list(alpha1=matrix(0, ntrans, nsegs), alpha0=0)}

params<-c("psi", "beta", "p", "alpha0", "psihat", "thetamean", "deltamean",
  "deltastarmean", "uhat")

out<-bugs(data, inits, params, "model.txt", n.chains=3,
n.iter=5000, n.burn=500, n.thin=10, debug=TRUE)

print(out, dig=3)


```



